Revealing transient strain in geodetic data with Gaussian process regression

### 2017-11-10 ResponseToReviewers/ResponseToReviewers.tex

\begin{document}

\signature{Trever T. Hines}

\begin{letter}{}
\opening{Dear Editor,}

% NOTE THE FOLLOWING ERRORS

% STRAIN COVARIANCE WAS INCORRECT

% THE NOISE MODEL WAS INCORRECT BECAUSE I USED A DIRAC RATHER THAN A
% KRONECKER DELTA.

@@

I believe the reviewer is refering to page 4-5 of the original draft,
where I first introduce $\eta$. I agree that a more thorough
discussion on the noise model is necessary, but I do not think this is
an appropriate place in the text for it. The paper is laid out so that
Section 2 contains a general description of the method for estimating
transient strain rates, and Section 3 contains details that are
specific to the particular application. There is no universally
appropriate noise model, so I decided to save a discussion on the
noise model for Section 3. This is clarified in Section 2 of the
revised draft when I say "The appropriate noise model may vary
depending on the application, and we hold off on specifying the
covariance matrix, $C_{\eta_i}$, until Section 3.1."

\textbf{Now, I am confused When looking at p.8 Section 4.1, the
description of the noise models should be given as a subsection of
Section 2 p.4. I would also improve the literature review. Several
other models have also been discussed (i.e. Montillet et al., 2014
uses a fractional Brownian motion model; ARMA, ARFIMA, GGM or Band
pass noise – see Het et al., 2017).}

Again, we think that the description of the noise model should remain
in Section 3 because it is specific to the Pacific Northwest
application.

We have added a few more references in Section 3.1 of the revised
draft to emphasize that GNSS noise modeling is a well-worked area. We
refer the reader to Bock and Melgar 2016 and He et al 2017 for a more
complete review.

@@

\textbf{Reviewer \#2:}\newline

\textbf{Comments to the Author(s)}
\textbf{Hines/Hetland GJI Review 08/2017}

\textbf{This is a well written paper and I have just a few minor
clarifications and suggestions below.  One point that might be of
interest to address is would it be possible to “approximate” this
approach so that it can run much faster?  I gather from the text that
the method is intensive for CPU and memory use.}

Our approach is indeed computationally intensive and approximation
methods for GPR do exist. In the discussion section for the revised
draft, we suggest exploring these approximation methods in future
research.

@@

\textbf{Page 3: Last line: min(t,t’) needs as scaling factor to make
units for covariance e.g., m$^2$/yr}

We added the scale factor $\phi$ to the covariance function for
Brownian motion.

@@

\textbf{Page 4: Stochastic model: It would seem that the assumption of no
covariance between the north and east components is not a very good
one since there is a tectonic framework in which the transients occur
and its unlikely this framework aligns along the Cardinal directions.
Maybe add a statement of impact of the neglect and possibly the idea
of reorienting the “axes” to align with the tectonic framework e.g.,
perpendicular to the subduction zone interface in this case?}

This is a good point that we neglected to mention in the previous
draft. We have added a paragraph to Section 2 that explains the
reasoning and implications for this assumption. The paragraph starts
with "It should be noted that we have ignored any covariances ...".

@@

\textbf{Page 4/5 Equation (3): Is it worth discussing at this point whether
the estimates of u and n (eta) can be separated without explicit prior
knowledge (which we may not have).  For example, orbit modeling errors
on one satellite (e.g., due to an unknown yaw problem in the
spacecraft) wilt generate a spatially and temporally correlated error
in $d_{ij}$.  How does this (stochastically non-modeled) error not project
in the u estimates?}

TODO

\textbf{Page 6: Equation 10: Maybe it is discussed later in the paper but it
is probably worthwhile stating at this point how the inverse is
performed with a block 0 in the matrix?}

The matrix being inverted in eq. 10 and 11 of the revised draft is
still invertible with the block 0, provided that the columns of
$\mitbf{G}$ are linearly dependent. This is explained in the revised
draft.

@@

\textbf{Page 6: Equation 12: Maybe some additional explanation is needed here.
How are these partial derivatives formed.  My understanding the u
estimates at this point correspond one-to-one with the positions and
times of the GNSS position determinations.  Did I miss something?}

The posterior displacements, $\hat{u}$, are actually a spatially and
temporally continuous stochastic process. We can evaluated $\hat{u}$
at any position and time that we may be interested in. We are also
able to compute the spatial and temporal derivatives analytically. In
Section 2 of the revised draft, we explain how the derivatives are
computed in greater detail. We hope this clarifies any confusion.

@@

\textbf{Page 8: Partly addresses issue raised about eqn 3.  It
probably should be noted that wether conditions east of -121 deg
longitude are different to the coast so there could be possible
problems with this noise model.}

This is also a good point that we neglected to mention in the first
draft. In Section 3.1 of the revised draft, we acknowledge the
potentially dubious assumption that noise east of -121 is
representative of the entire region. We added the disclaimer saying
"We assume the noise at these inland stations is representative of the
noise at all the stations considered in this study, which is probably
a poor assumption since the inland stations are subject to distincly
different climatic conditions".

### 2017-11-10 RevisedSubmission/hines_hetland_2017_gji.tex
did more

We let $\vec{u}(p) = (u_\e(p),u_\n(p))$ be our prior understanding of
transient displacements. Since transient displacements are not
precisely known, we cannot consider $\vec{u}$ to be a deterministic
function. Instead, $\vec{u}$ is considered to be a stochastic process
containing a distribution of functions that could potentially describe
transient displacements. Specifically, we let each component of
$\vec{u}$ be a Gaussian process. A Gaussian process is a stochastic
process whose value at any collection of points can be described with
a multivariate normal distribution. That is to say, the random vector
$[u_i(p)]_{p \in \points}$ has a multivariate normal distribution for
any collection of points $\points$. A realization of the random vector
$[u_i(p)]_{p \in \points}$ can be interpretted as a realization of
$u_i$ (i.e., a sample function) that is evaluated at $\points$. Just
as a multivariate normal distribution is fully determined by a mean
vector and a covariance matrix, a Gaussian process is fully determined
by a mean function and a covariance function. For example, Brownian
motion, $B(t)$, is a well known Gaussian process in $\mathbb{R}^1$
which has the mean function $\E{B(t)} = 0$ and the covariance function
$\Cov{B(t),B(t')} = \min(t,t')$, where $t,t' \ge 0$. We let each
component of $\vec{u}$ have zero mean and a generic covariance
function $\Cov{u_i(p),u_i(p')} = C_{u_i}(p,p')$. Using a more concise
notation, we write our prior on each component of $\vec{u}$ as $u_i
\sim \mathcal{GP}\left(0,C_{u_i}\right)$.

@@

The transient strain rates shown for the winter 2015-2016 SSE in
Figure \ref{fig:StrainMap} are characteristic of most SSEs in the
Puget Sound region. The SSEs cause trench perpendicular compression in
the Olympic Peninsula and extension east of Puget Sound. The strain
transitions from compression to extension around the southern tip of
Vancouver Island, coinciding with where fault slip tends to be
inferred \citep[e.g.,][]{Dragert2001, Wech2009, Schmidt2010}. Thus,
this pattern of strain is to be expected. Over decadal timescales,
there is trench perpendicular compression throughout this study region
caused by steady tectonic plate motion \citep{Murray2000,
McCaffrey2007, McCaffrey2013}. When comparing the transient strain
rates caused by SSEs to the secular strain rates, we see that SSEs are
concentrating tectonically accumulated strain energy towards the
trench, and presumably pushing the subduction zone closer to failure.

>>>>

We let $\vec{u}(p) = (u_\e(p),u_\n(p))$ be our prior understanding of
transient displacements. Since transient displacements are not
precisely known, we cannot consider $\vec{u}$ to be a deterministic
function. Instead, $\vec{u}$ is considered to be a stochastic process
containing a distribution of functions that could potentially describe
transient displacements. Specifically, we let each component of
$\vec{u}$ be a Gaussian process. A Gaussian process is a stochastic
process whose value at any collection of points can be described with
a multivariate normal distribution. That is to say, the random vector
$[u_i(p)]_{p \in \points}$ has a multivariate normal distribution for
any collection of points $\points$. A realization of the random vector
$[u_i(p)]_{p \in \points}$ can be interpretted as a realization of
$u_i$ (i.e., a sample function) that is evaluated at $\points$. Just
as a multivariate normal distribution is fully determined by a mean
vector and a covariance matrix, a Gaussian process is fully determined
by a mean function and a covariance function. For example, Brownian
motion, $B(t)$, is a well known Gaussian process in $\mathbb{R}^1$
which has the mean function $\E{B(t)} = 0$ and the covariance function
$\Cov{B(t),B(t')} = \phi^2 \min(t,t')$, where $t,t' \ge 0$ and $\phi$
is a scale factor. We let each component of $\vec{u}$ have zero
mean and a generic covariance function $\Cov{u_i(p),u_i(p')} =
C_{u_i}(p,p')$. Using a more concise notation, we write our prior on
each component of $\vec{u}$ as $u_i \sim
\mathcal{GP}\left(0,C_{u_i}\right)$.

@@

The transient strain rates shown for the winter 2015-2016 SSE in
Figure \ref{fig:StrainMap} are characteristic of most SSEs in the
Puget Sound region. The SSEs cause trench perpendicular compression in
the Olympic Peninsula and extension east of Puget Sound. The strain
transitions from compression to extension around the southern tip of
Vancouver Island, coinciding with where fault slip tends to be
inferred \citep[e.g.,][]{Dragert2001, Wech2009, Schmidt2010}. Thus,
this pattern of strain is to be expected. Over decadal time-scales,
there is trench perpendicular compression throughout this study region
caused by steady tectonic plate motion \citep{Murray2000,
McCaffrey2007, McCaffrey2013}. When comparing the transient strain
rates caused by SSEs to the secular strain rates, we see that SSEs are
concentrating tectonically accumulated strain energy towards the
trench, and presumably pushing the subduction zone closer to failure.
